---
format: html
editor-options: 
  chunk-output-type: console
---

# Analysis

## Setup

```{r}
#| label: libraries

library(rlang)
library(tidyverse)
tprint <- 75 # default tibble print
options(tibble.print_max = tprint, tibble.print_min = tprint) # show up to tprint rows

library(fs)

# tools
library(vroom)
library(readxl)
library(openxlsx) # for writing xlsx files
library(lubridate)
library(RColorBrewer)
library(RcppRoll)
library(fredr)
library(tidycensus)

# boyd libraries
library(btools)
library(bdata)
library(bggtools)
library(bmaps)

# graphics
library(scales)
library(ggbeeswarm)
library(patchwork)
library(gridExtra)
library(ggrepel)
library(ggbreak)

# tables
library(formattable)
library(knitr)
library(kableExtra)
library(DT)
library(gt)
library(gtExtras)
library(janitor)
library(skimr)
library(vtable)

# maps
# library(maps)
# # https://cran.r-project.org/web/packages/usmap/vignettes/mapping.html
# library(usmap)

source(here::here("R", "libraries_ts.R"))

```

```{r}
#| label: constants
# E:\data\nys_monthly_tax\data
# monthly_collections.rds

DRES <- here::here("results")

DDTF <- r"(E:\data\nys_monthly_tax\data)"

x90 <- theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust = 0.5))

```

## Get data

Note that units are DOLLARS (not thousands).

I think the nese tax was included in the state PIT for a while?? go back and check

```{r}
#| label: get-data

df <- readRDS(fs::path(DDTF, "monthly_collections.rds"))
glimpse(df)

count(df |> filter(vtype == "local"), vname, tax)

# df |>
#   filter(vname == "local_congestion") |>
#   ggplot(aes(date, value)) +
#   geom_line() +
#   scale_x_date(breaks = "1 year")

first_data_date <- "2009-10-01"
pmt <- df |>
  filter(str_detect(vname, "pmt"), date >= first_data_date) |>
  mutate(vname = str_remove(vname, "local_")) |>
  select(date, vname, tax, value)
glimpse(pmt)
count(pmt, vname, tax)

pmt2 <- pmt |>
  select(-tax) |>
  pivot_wider(names_from = vname, values_fill = 0) |>
  mutate(pmttot = pmtwage + pmtnesenet) |>
  select(
    date,
    pmtwage,
    pmtnesenet,
    pmttot,
    pmtnesegross,
    pmtscoffset,
    pmtneseref
  )

pmt2

pmtlong <- pmt2 |>
  select(date, pmtwage, pmtnesenet, pmttot) |>
  pivot_longer(-date, names_to = "vname")

pmt2 |> tail(12)

```

Construct state pit approximating the NESE gross concept:

PIT estimated + final

```{r}
#| label: state-pit

count(df |> filter(vtype == "pit"), vname, tax)

nys <- df |>
  filter(vname %in% c("pit_wh", "pit_et", "pit_fr", "pit_ref")) |>
  filter(year(date) >= 2019) |>
  select(date, vname, value) |>
  pivot_wider(names_from = vname) |>
  mutate(pitetfr = pit_et + pit_fr)
nys

pmt2 |>
  ggplot(aes(date, pmtscoffset)) +
  geom_line() +
  geom_point()


```

```{r}
#| label: show-tax
#| output: true

# pmtlong |>
#   ggplot(aes(date, value)) +
#   geom_line() +
#   facet_wrap(~vname, scales = "free")

pmtlong |>
  filter(vname == "pmtwage") |>
  filter(year(date) >= 2019) |>
  mutate(date = yearmonth(date)) |>
  as_tsibble(index = date) |>
  feasts::gg_season(value) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    name = "$ millions",
    labels = label_comma(scale = 1e-3, accuracy = 1)
  ) +
  labs(x = "") +
  ggtitle("PMT wage tax revenue") +
  theme_bw() +
  x90

pmtlong |>
  filter(vname == "pmtnesenet") |>
  filter(year(date) >= 2019) |>
  mutate(date = yearmonth(date)) |>
  as_tsibble(index = date) |>
  feasts::gg_season(value) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    name = "$ millions",
    labels = label_comma(scale = 1e-3, accuracy = 1)
  ) +
  labs(x = "") +
  ggtitle("PMT NESE payments") +
  theme_bw() +
  x90

```

## MTA NESE tax compared to state estimated tax and final returns

What is going on with MTA October NESE collections in 2022, 2023, 2024???

NESE is on the same payment schedule as state PIT estimated payments: Apr, Jun, Sep, and Jan. So why is the MTA 2022, 2023, and 2024 bump in Oct not Sep? Given that it starts in 2022, it probably (?) is not related to the 2023 tax increase. Therefore it might not be something we should think of as related to prior years -- it could be current money.

```{r}
#| label: nesenet-vs-state

pdata <- bind_rows(
  pmt2 |>
    filter(year(date) >= 2019) |>
    select(date, value = pmtnesenet) |>
    mutate(type = "mta"),
  nys |>
    filter(year(date) >= 2019) |>
    select(date, value = pitetfr) |> # pit_et or pitetfr
    mutate(type = "nys")
)

p <- pdata |>
  filter(year(date) >= 2019) |>
  mutate(
    year = year(date) |> as.factor(),
    monthf = factor(month.abb[month(date)], levels = month.abb)
  ) |>
  ggplot(aes(monthf, value, colour = year, group = year)) +
  geom_line() +
  geom_point() +
  scale_x_discrete(name = NULL, limits = month.abb) +
  scale_y_continuous(
    name = "Collections, $ millions",
    labels = scales::label_comma(scale = 1e-6)
  ) +
  facet_wrap(facets = ~type, ncol = 1, scales = "free") +
  theme_bw() +
  ggtitle(
    label = "MTA Net earnings from self-employment (NESE) tax compared to NYS PIT estimated tax + final returns"
  )
p

ggsave(filename = fs::path(DRES, "nesenet_vs_nys.png"), plot = p, width = 10)

```


```{r}
#| label: nesegross-vs-state

pdata <- bind_rows(
  pmt2 |>
    filter(year(date) >= 2019) |>
    select(date, value = pmtnesegross) |>
    mutate(type = "mta"),
  nys |>
    filter(year(date) >= 2019) |>
    select(date, value = pitetfr) |> # pit_et or pitetfr
    mutate(type = "nys")
)

p <- pdata |>
  filter(year(date) >= 2019) |>
  mutate(
    year = year(date) |> as.factor(),
    monthf = factor(month.abb[month(date)], levels = month.abb)
  ) |>
  ggplot(aes(monthf, value, colour = year, group = year)) +
  geom_line() +
  geom_point() +
  scale_x_discrete(name = NULL, limits = month.abb) +
  scale_y_continuous(
    name = "Collections, $ millions",
    labels = scales::label_comma(scale = 1e-6)
  ) +
  facet_wrap(facets = ~type, ncol = 1, scales = "free") +
  theme_bw() +
  ggtitle(
    label = "MTA Net earnings from self-employment (NESE) tax GROSS compared to NYS PIT estimated tax + final returns"
  )
p

ggsave(filename = fs::path(DRES, "nesegross_vs_nys.png"), plot = p, width = 10)

```



```{r}
#| label: nese-only

pdata <- pmt2 |>
  filter(year(date) >= 2022) |>
  select(date, pmtnesenet, pmtnesegross) |>
  pivot_longer(-date)

p <- pdata |>
  ggplot(aes(date, value, colour = name)) +
  geom_line() +
  geom_point() +
  scale_x_date(
    name = NULL,
    breaks = "3 months",
    labels = scales::label_date(format = "%Y-%b")
  ) +
  scale_y_continuous(
    name = "Collections, $ millions",
    labels = scales::label_comma(scale = 1e-6)
  ) +
  x90 +
  theme_bw() +
  ggtitle(
    label = "MTA Net earnings from self-employment (NESE) tax"
  )
p


pdata <- pmt2 |>
  filter(year(date) >= 2022) |>
  select(date, pmtscoffset) |>
  pivot_longer(-date)

p <- pdata |>
  ggplot(aes(date, value, colour = name)) +
  geom_line() +
  geom_point() +
  scale_x_date(
    name = NULL,
    breaks = "3 months",
    labels = scales::label_date(format = "%Y-%b")
  ) +
  scale_y_continuous(
    name = "Collections, $ millions",
    labels = scales::label_comma(scale = 1e-6)
  ) +
  x90 +
  theme_bw() +
  ggtitle(
    label = "MTA Net earnings from self-employment (NESE) tax"
  )
p

# ggsave(filename = fs::path(DRES, "nese_vs_nys.png"), plot = p, width = 10)
```

## MTA wage tax compared to state withholding tax

```{r}
#| label: wagetax-vs-state

pdata <- bind_rows(
  pmt2 |>
    filter(year(date) >= 2019) |>
    select(date, value = pmtwage) |>
    mutate(type = "mta"),
  nys |>
    filter(year(date) >= 2019) |>
    select(date, value = pit_wh) |>
    mutate(type = "nys")
)

p <- pdata |>
  filter(year(date) >= 2019) |>
  mutate(
    year = year(date) |> as.factor(),
    monthf = factor(month.abb[month(date)], levels = month.abb)
  ) |>
  ggplot(aes(monthf, value, colour = year, group = year)) +
  geom_line() +
  geom_point() +
  scale_x_discrete(name = NULL, limits = month.abb) +
  scale_y_continuous(
    name = "Collections, $ millions",
    labels = scales::label_comma(scale = 1e-6)
  ) +
  facet_wrap(facets = ~type, ncol = 1, scales = "free") +
  theme_bw() +
  ggtitle(
    label = "MTA PMT wage tax compared to NYS PIT withholding"
  )
p

ggsave(filename = fs::path(DRES, "wagetax_vs_nys.png"), plot = p, width = 10)

p <- pdata |>
  arrange(type, date) |>
  mutate(pch = value / lag(value, 12) - 1, .by = type) |>
  filter(year(date) >= 2024) |>
  ggplot(aes(date, pch, colour = type)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_date(
    name = NULL,
    breaks = "3 months",
    labels = scales::label_date(format = "%Y-%b")
  ) +
  scale_y_continuous(
    breaks = seq(-.1, 1, .05),
    labels = scales::label_percent(accuracy = 1)
  ) +
  ggtitle(
    label = "MTA wage tax and NYS withholding tax",
    subtitle = "% growth year over year"
  ) +
  theme_bw()
p

ggsave(
  filename = fs::path(DRES, "wagetax_pch_vs_nys.png"),
  plot = p,
  width = 10
)

```
